---
title: "10a_PoEdata"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup_poe, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

In this tenth session, we will explore how to access and use econometrics datasets for hands-on practice using R. The datasets used in this session are sourced from the *Principles of Econometrics, 4th Edition* by Hill, Griffiths, and Lim. They are available in the following formats:

* **Stata (.dta)**:
  [http://www.principlesofeconometrics.com/poe4/poe4stata.htm](http://www.principlesofeconometrics.com/poe4/poe4stata.htm)

* **Excel Spreadsheet**:
  [http://www.principlesofeconometrics.com/poe4/poe4excel.htm](http://www.principlesofeconometrics.com/poe4/poe4excel.htm)

* **R Dataset Package**:
  [https://github.com/ccolonescu/PoEdata](https://github.com/ccolonescu/PoEdata)

Although the original format is intended for Stata, this course will use R for all analytical work.

## 2. Installing the `PoEdata` Package

The `PoEdata` package contains a collection of datasets for econometric practice. It was developed by **Constantin Colonescu** and made available on GitHub.

### 2.1 Step-by-Step Installation

```{r install_PoEdata, eval=FALSE}
## Install devtools if not already installed
install.packages("devtools")

## Load devtools
library(devtools)

## Install PoEdata from GitHub
install_git("https://github.com/ccolonescu/PoEdata")
```

### 2.2 Common Installation Issues and Solutions

#### 2.2.1 Rtools Not Installed

**Error Message:**

```
WARNING: Rtools is required to build R packages, but is not currently installed.
```

**Solution:**

* Download and install Rtools 4.5:
  [https://cran.r-project.org/bin/windows/Rtools/](https://cran.r-project.org/bin/windows/Rtools/)
* During installation, check the option to **add Rtools to system PATH**.
* Verify installation:

```{r, eval=FALSE}
pkgbuild::has_build_tools(debug = TRUE)
```

#### 2.2.2 Git Not Installed

**Error Message:**

```
Error: Failed to install 'unknown package' from Git:
  Git does not seem to be installed on your system.
```

**Solution:**

* Download Git for Windows:
  [https://git-scm.com/download/win](https://git-scm.com/download/win)
* Choose **"Git from the command line and also from 3rd-party software"** during installation.
* Restart RStudio and verify:

```{r, eval=FALSE}
system("git --version")
```

Expected output:

```
git version 2.43.0.windows.1
```

#### 2.2.3 Alternative Installation via `remotes` (Recommended)

```{r, eval=FALSE}
install.packages("remotes")  ## if not already installed
remotes::install_github("ccolonescu/PoEdata")
```

### 2.3 Troubleshooting PATH Issues

If R returns `[1] 127` after Git installation, it means Git is not found in system PATH. Fix by:

1. Locate Git path (e.g., `C:\\Program Files\\Git\\cmd`)
2. Open:
   **Control Panel → System → Advanced System Settings → Environment Variables**
3. Edit the `Path` variable and add Git directory
4. Restart RStudio

Then test again:

```{r, eval=FALSE}
system("git --version")
```

## 3. Summary of Installation

By following the steps above, you will be able to:

* Install essential tools (`Git`, `Rtools`, `PoEdata`)
* Resolve common installation errors
* Prepare your R environment for hands-on econometric analysis

```{r}
library(PoEdata)
```

## 4. Food Expenditure Data

### 4.1 Load and Examine the Data

```{r}
data(food)
str(food)
head(food, 5)
food[1:5, "food_exp"]
subset(food, income < 10)
```

### 4.2 Summary Statistics

```{r}
summary(food)
summary(food$food_exp)
```

### 4.3 Scatter Plot: Food Expenditure vs Income

```{r}
library(ggplot2)

ggplot(food, aes(x = income, y = food_exp)) +
  geom_point() +
  labs(title = "Food Expenditure Data",
       x = "Income", y = "Food Expenditure") +
  scale_y_continuous(breaks = seq(0, 600, 100)) +
  scale_x_continuous(breaks = seq(0, 35, 5))
```

### 4.4 Linear Regression and Elasticity

```{r}
model1 <- lm(food_exp ~ income, data = food)
summary(model1)

food$yhat <- predict(model1)
food$ehat <- residuals(model1)

mean_income <- mean(food$income)
mean_yhat <- mean(food$yhat)
elas_mean <- coef(model1)["income"] * mean_income / mean_yhat

food$elas <- coef(model1)["income"] * food$income / food$yhat
summary(food$elas)
```

### 4.5 Plot with Fitted Line

```{r}
ggplot(food, aes(x = income, y = food_exp)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Fitted Regression Line",
       x = "Income", y = "Food Expenditure") +
  scale_y_continuous(breaks = seq(0, 600, 100)) +
  scale_x_continuous(breaks = seq(0, 35, 5))
```

## 5. Quadratic Model Example

### 5.1 Load and Estimate Model

```{r}
data(br)
summary(br)

br$sqft2 <- br$sqft^2
model2 <- lm(price ~ sqft2, data = br)
br$priceq <- predict(model2)
```

### 5.2 Plot Quadratic Fit

```{r}
ggplot(br, aes(x = sqft, y = price)) +
  geom_point() +
  geom_line(aes(y = priceq), color = "blue", linewidth = 1) +
  labs(title = "Quadratic Fit", x = "Square Foot", y = "Price")
```

### 5.3 Slope and Elasticity at a Point

```{r}
b2 <- coef(model2)["sqft2"]
intercept <- coef(model2)["(Intercept)"]

slope_2000 <- 2 * b2 * 2000
pred_2000 <- intercept + b2 * 2000^2
elas_2000 <- slope_2000 * 2000 / pred_2000
```

## 6. Log-linear Model

```{r}
br$lprice <- log(br$price)

hist(br$price, breaks = 20, main = "Price Distribution", xlab = "Price", freq = FALSE)
hist(br$lprice, breaks = 20, main = "Log Price Distribution", xlab = "Log(Price)", freq = FALSE)

model3 <- lm(lprice ~ sqft, data = br)
br$lpricef <- predict(model3)
br$pricef <- exp(br$lpricef)

ggplot(br, aes(x = sqft, y = price)) +
  geom_point() +
  geom_line(aes(y = pricef), color = "red", linewidth = 1)
```

### 6.1 Elasticity and Marginal Effects

```{r}
b_sqft <- coef(model3)["sqft"]
br$me <- b_sqft * br$pricef
br$elas <- b_sqft * br$sqft

summary(br$me)
summary(br$elas)
```

## 7. Indicator Variable Model (University Town)

```{r}
data(utown)
summary(utown)

ggplot(utown, aes(x = price)) +
  geom_histogram(data = subset(utown, utown == 0), binwidth = 12, fill = "blue", alpha = 0.5) +
  geom_histogram(data = subset(utown, utown == 1), binwidth = 12, fill = "green", alpha = 0.5) +
  facet_wrap(~ utown, ncol = 1) +
  labs(title = "House Prices by Neighborhood", x = "Price ($1000)")

model4 <- lm(price ~ factor(utown), data = utown)
summary(model4)

t.test(price ~ utown, data = utown)
```

## 8. Simulated Data Experiment

### 8.1 Generate and Model Simulated Data

```{r}
set.seed(1234567)
n <- 40
beta1 <- 100
beta2 <- 10
sigma <- 50

x <- rep(c(10, 20), each = n/2)
e <- rnorm(n, mean = 0, sd = sigma)
y <- beta1 + beta2 * x + e

sim_data <- data.frame(x, y)
model_sim <- lm(y ~ x, data = sim_data)
summary(model_sim)
```

### 8.2 Monte Carlo Simulation

```{r}
simulate_model <- function() {
  x <- rep(c(10, 20), each = n/2)
  e <- rnorm(n, mean = 0, sd = sigma)
  y <- beta1 + beta2 * x + e
  model <- lm(y ~ x)
  return(c(b1 = coef(model)[1], b2 = coef(model)[2], sig2 = sigma^2))
}

sim_results <- replicate(1000, simulate_model())
sim_df <- as.data.frame(t(sim_results))

summary(sim_df)
hist(sim_df$b2, freq = FALSE, main = "Sampling Distribution of Slope", xlab = "Slope Estimates")
```

